A Survey of Real-Time Strategy Game AI\\ Research and Competition in StarCraft

### 2013-03-16 survey.tex
figures updated
incorporated Dave's feedback

\begin{itemize}
\item They are {\em simultaneous move} games, where more than one player can issue actions at the same time. Additionally, these actions are {\em durative}, i.e. actions are not instantaneous, but take some amount of time time to complete.
\item RTS games are ``real-time'', which actually means is that each player has a very small amount of time to decide the next move. Compared to Chess, where players have several minutes to decide the next action, in StarCraft, the game executes at 24 iterations per second, which means that players can act as fast as every 42ms, before the game state changes.
\item Most RTS games are partially observable: players can only see the part of the map that has been explored. This is referred to as the {\em fog-of-war}.
\item Most RTS games are non-deterministic. Some actions have a chance of success.
\item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large. For example, the state space of Chess is typically estimated to be around $10^{50}$, heads up no-limit texas holdem poker around $10^{80}$, and Go around $10^{170}$. In comparison, the state space of StarCraft in a typical map is estimated to be many orders of magnitude larger than any of those, as discussed in the next section.
\end{itemize}

@@

{\em StarCraft: Brood War} is an immensely popular RTS game released in 1998 by Blizzard Entertainment. StarCraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg. One of the most remarkable aspects of StarCraft is that the three races are extremely well balanced:

@@

Figure~\ref{fig:StarCraft} shows a screenshot of StarCraft showing a player playing Terrans. In order to win a StarCraft game, players must first gather resources (minerals and Vespene gas). As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units), research new technologies (in order to use new unit abilities or improve the units) and train attack units. Units must be distributed to accomplish different tasks such as reconnaissance, defense and attack. While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts. Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight a battle, which requires quick and reactive control of each of the units.

@@

A typical StarCraft map is defined as a rectangular grid, where the $width \times height$ of the map is measured in the number of $32 \times 32$ square of pixels, also knows as build tiles. Although the
resolution  at which  units  move  is in squares of $8 \times 8$ pixels, also knows as walk tiles. The typical dimensions for maps range from $64 \times 64$ to $256 \times 256$ build tiles. Each player  can
control up  to 200 units (not including  buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  each  one  with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. For instance, in the game ladder iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked by their current point totals ($E$ being the lowest possible rank, and $A^+$ and $Olympic$ being the second highest and highest ranks, respectively), the best StarCraft AI bots are ranked between $D$ and $D^+$, where average  amateur players are  ranked between $C^+$ and  $B$. For comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

@@

\subsubsection{Spatial and Temporal Reasoning}
% TODO just a paragraph on exposing why there are temporal and spatial
% reasoning in RTS AI.
Spatial  reasoning   is  related  to   each  aspect  of   terrain
exploitation.  Spatial reasoning is involved in tasks such as building placing, or base expansion. In the former, the player needs to
carefully  consider building  position  into its  own  bases to  both
protect them  by creating  a wall against  invasions and to  avoid bad
configurations  where large  units could  be stuck. In base expansion, the player has to
choose good available emplacements to  build a new base, regarding its
own  position  and opponent's  bases.  Finally,  spatial reasoning  is
key to tactical reasoning: players need to decide where to place units
for battle,  favoring, for  instance, engagements when  the opponent's
units are into a bottleneck.

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
% [Alberto: I guess Weber's number comes from StarCraft walkthroughs like this: http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt, there is also a chance to hit in even grounds (like 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
For instance, in StarCraft, it is always an advantage to  have its units  on high ground  whereas enemy is  on low ground, since units  on low ground have no vision  on high ground, and firing from low ground have a chance to  hit units on high ground.

@@

\begin{itemize}
\item
  BroodwarBotQ\footnote{\url{http://github.com/SnippyHolloW/BroodwarBotQ}}:
  uses  abstraction for  {\em combat}, and  divide-and-conquer for
  {\em economy}  and  {\em intelligence  gathering}. To  avoid  conflicts
  between modules (since  the individual tasks of each  of the modules
  are not completely independent), BBQ uses an arbitrator.
\item  Nova\footnote{\url{http://nova.wolfwork.com/}}:  is similar  in
  design as  BroodwarBotQ, and uses  abstraction for {\em combat},
  and  divide-and-conquer for  {\em economy}.  The differences  are
  that Nova does not have  an arbitrator to resolve conflicts, but has
  a  higher-level   module  ({\em  strategy   manager}),  which  posts
  information to the blackboard that the rest of modules follow (thus,
  making use of abstraction).
\item
  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
  also  uses abstraction  in  {\em combat} like  the previous  two
  bots. But it  also uses it in {|em economy}: as  can be seen, the
  production manager sends commands to the building manager, who is in
  charge   of   producing   the   buildings.  This   bot   also   uses
  divide-and-conquer, and  tasks like scouting  and resource gathering
  are managed by separate, independent modules.
\item       Skynet\footnote{\url{http://code.google.com/p/skynetbot/}}:
  makes     extensive     use      of     both     abstraction     and
  divide-and-conquer.  We  can see a  high level module  that issues commands  to a
  series of  tactics modules. The  collection of tactic  modules queue
  {\em  tasks} (that  are analogous  to the  abstract actions  used in
  SPAR).  Each different  task has  a specific  low level  module that
  knows how to  execute it. Thus, Skynet uses  a 3 layered abstraction
  hierarchy,  and uses  divide-and-conquer  in all  levels except  the
  highest.
\item
  SPAR\footnote{\url{http://www.planiart.usherbrooke.ca/projects/spar/}}:
  only uses abstraction. Its high-level module determines the strategy
  to  use,  and  the  tactical  decision  module  divides  it  into  a
  collection  of {\em  abstract  actions}, that  are  executed by  the
  lower-level modules.
\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
  mainly divide-and-conquer oriented,
  with a  slight abstraction on  {\em economy} due to a  SpendManager deciding
  how  to  spend  and  share  resources  among  Base,  Production  and
  Construction Managers.  At the beginning of a  game, the MoodManager
  initializes  a  ``mood''  which  will  influence  both  tactics  and
  strategy. {\em Combat} is divided into three  independent managers: the
  {\em  Defense Manager},  controlling  military units  when there  is
  nothing special, the {\em  Under Attack Manager}, activated when the
  opponent is attacking our bases,  and the {\em Army Manager}, taking
  control of units when it is time to attack, following a timing given
  by the  current mood. This bot  does not manage however  any kind of
  reactive controls so far.
\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
  two-tier  abstraction hierarchy,  where a  collection  of high-level
  modules command a collection of lower-level agents in charge of each
  of  the units.  At  the high-level,  BTHAI uses  divide-and-conquer,
  having   multiple  high-level  modules   issuing  commands   to  the
  lower-level units.
\end{itemize}

@@

\begin{figure*}[t]
	%\vspace*{-2ex}
    \centering
    \includegraphics[width=0.34\textwidth]{figures/vstable3.pdf}
    \includegraphics[width=0.3\textwidth]{figures/vstable6.pdf}
    \includegraphics[width=0.3\textwidth]{figures/vstable.pdf}
	%\vspace*{-4ex}
    \caption{Win percentages of CIG 2012 competition, from left
    to right: 3-player maps only, 6-player maps only, all maps.
    Read from line to column, bot in row wins given fraction of
    games against bot in column. For some bots, we find interesting
    differences, e.g. Xelnaga gets worse on 6-player maps, UAlbertaBot
    gets better. Only Xelnaga can reliably beat Skynet, but only on 
    3-player maps}
    \label{fig:cig2012-mapresults}
\end{figure*}

@@

\subsection{StarCraft Bot Ladder}\label{sec:cig2011}

@@

As argued before, addressing these open research questions would not just lead to better RTS game playing bots, but would provide means to improve many other complex AI components, not solely focussing on games. This would apply to many areas that are in need of real-time decision components working under uncertainty in the form of incomplete and/or noisy information, such as  e.g. robotics and assistance systems.
%bring significant advances to many other areas of artificial intelligence. This is so because RTS games embody many characteristics of many other problems appearing in many areas of AI (such as robotics, multiagent systems, etc.).

>>>>

\begin{itemize}
\item They are {\em simultaneous move} games, where more than one player can issue actions at the same time. Additionally, these actions are {\em durative}, i.e. actions are not instantaneous, but take some amount of time time to complete.
\item RTS games are ``real-time'', which actually means is that each player has a very small amount of time to decide the next move. Compared to Chess, where players may have several minutes to decide the next action, in StarCraft, the game executes at 24 iterations per second, which means that players can act as fast as every 42ms, before the game state changes.
\item Most RTS games are partially observable: players can only see the part of the map that has been explored. This is referred to as the {\em fog-of-war}.
\item Most RTS games are non-deterministic. Some actions have a chance of success.
\item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large. For example, the state space of Chess is typically estimated to be around $10^{50}$, heads up no-limit texas holdem poker around $10^{80}$, and Go around $10^{170}$. In comparison, the state space of StarCraft in a typical map is estimated to be many orders of magnitude larger than any of those, as discussed in the next section.
\end{itemize}

@@

{\em StarCraft: Brood War} is an immensely popular RTS game released in 1998 by Blizzard Entertainment. StarCraft is set in a science-fiction based universe where the player must choose one of the three races: Terran, Protoss or Zerg. One of the most remarkable aspects of StarCraft is that the three races are extremely well balanced:

@@

Figure~\ref{fig:StarCraft} shows a screenshot of StarCraft showing a player playing the Terran race. In order to win a StarCraft game, players must first gather resources (minerals and Vespene gas). As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units), research new technologies (in order to use new unit abilities or improve the units) and train attack units. Units must be distributed to accomplish different tasks such as reconnaissance, defense and attack. While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts. Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight a battle, which requires quick and reactive control of each of the units.

@@

A typical StarCraft map is defined as a rectangular grid, where the $width \times height$ of the map is measured in the number of $32 \times 32$ square of pixels, also knows as build tiles. Although the
resolution  at which  units  move  is in squares of $8 \times 8$ pixels, also knows as walk tiles. The typical dimensions for maps range from $64 \times 64$ to $256 \times 256$ build tiles. Each player  can
control up  to 200 units (plus an unlimited number of buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  each  one  with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. For instance, in the game ladder iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked by their current point totals ($E$ being the lowest possible rank, and $A^+$ and $Olympic$ being the second highest and highest ranks, respectively), the best StarCraft AI bots are ranked between $D$ and $D^+$, where average  amateur players are  ranked between $C^+$ and  $B$. For comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

@@

\subsubsection{Spatial and Temporal Reasoning}
% TODO just a paragraph on exposing why there are temporal and spatial
% reasoning in RTS AI.
Spatial  reasoning   is  related  to   each  aspect  of   terrain
exploitation.  Spatial reasoning is involved in tasks such as building placing, or base expansion. In the former, the player needs to
carefully  consider building  position  into its  own  bases to  both
protect them  by creating  a wall against  invasions and to  avoid bad
configurations  where large  units could  be stuck. In base expansion, the player has to
choose good available locations to  build a new base, regarding its
own  position  and opponent's  bases.  Finally,  spatial reasoning  is
key to tactical reasoning: players need to decide where to place units
for battle,  favoring, for  instance, engagements when  the opponent's
units are into a bottleneck.

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
% [Alberto: I guess Weber's number comes from StarCraft walkthroughs like this: http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt, there is also a chance to hit in even grounds (like 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
Another example of spatial reasoning in StarCraft is that it is always an advantage to  have its units  on high ground  while enemy is on low ground, since units  on low ground have no vision  on high ground, and firing from low ground have a chance to  hit units on high ground.

@@

\begin{itemize}
\item
  BroodwarBotQ\footnote{\url{http://github.com/SnippyHolloW/BroodwarBotQ}}:
  uses  abstraction for  {\em combat}, and  divide-and-conquer for
  {\em economy}  and  {\em intelligence  gathering}. To  avoid  conflicts
  between modules (since  the individual tasks of each  of the modules
  are not completely independent), BBQ uses an arbitrator.
\item  Nova\footnote{\url{http://nova.wolfwork.com/}}:  is similar  in
  design as  BroodwarBotQ, and uses  abstraction for {\em combat},
  and  divide-and-conquer for  {\em economy}.  The differences  are
  that Nova does not have  an arbitrator to resolve conflicts, but has
  a  higher-level   module  ({\em  strategy   manager}),  which  posts
  information to the blackboard that the rest of modules follow (thus,
  making use of abstraction).
\item
  UAlbertaBot\footnote{\url{http://code.google.com/p/ualbertabot/}}:
  also  uses abstraction  in  {\em combat} like  the previous  two
  bots. But it  also uses it in {\em economy}: as  can be seen, the
  production manager sends commands to the building manager, who is in
  charge   of   producing   the   buildings.  This   bot   also   uses
  divide-and-conquer, and  tasks like scouting  and resource gathering
  are managed by separate, independent modules.
\item       Skynet\footnote{\url{http://code.google.com/p/skynetbot/}}:
  makes     extensive     use      of     both     abstraction     and
  divide-and-conquer.  We  can see a  high level module  that issues commands  to a
  series of  tactics modules. The  collection of tactic  modules queue
  {\em  tasks} (that  are analogous  to the  abstract actions  used in
  SPAR).  Each different  task has  a specific  low level  module that
  knows how to  execute it. Thus, Skynet uses  a 3 layered abstraction
  hierarchy,  and uses  divide-and-conquer  in all  levels except  the
  highest.
\item
  SPAR\footnote{\url{http://www.planiart.usherbrooke.ca/projects/spar/}}:
  only uses abstraction. Its high-level module determines the strategy
  to  use,  and  the  tactical  decision  module  divides  it  into  a
  collection  of {\em  abstract  actions}, that  are  executed by  the
  lower-level modules.
\item  AIUR\footnote{\url{http://code.google.com/p/aiurproject/}}:  is
  mainly divide-and-conquer oriented,
  with a  slight abstraction on  {\em economy} due to a  SpendManager deciding
  how  to  spend  and  share  resources  among  Base,  Production  and
  Construction Managers.  At the beginning of a  game, the MoodManager
  initializes  a  ``mood''  which  will  influence  both  tactics  and
  strategy. {\em Combat} is divided into three  independent managers: the
  {\em  Defense Manager},  controlling  military units  when there  is
  nothing special, the {\em  Under Attack Manager}, activated when the
  opponent is attacking our bases,  and the {\em Army Manager}, taking
  control of units when it is time to attack, following a timing given
  by the  current mood. This bot  does not manage however  any kind of
  reactive controls so far.
\item  BTHAI\footnote{\url{http://code.google.com/p/bthai/}}:  uses  a
  two-tier  abstraction hierarchy,  where a  collection  of high-level
  modules command a collection of lower-level agents in charge of each
  of  the units.  At  the high-level,  BTHAI uses  divide-and-conquer,
  having   multiple  high-level  modules   issuing  commands   to  the
  lower-level units.
\end{itemize}

@@

\begin{figure*}[t]
	%\vspace*{-2ex}
    \centering
    \includegraphics[width=0.32\textwidth]{figures/vstable3.pdf}
    \includegraphics[width=0.32\textwidth]{figures/vstable6.pdf}
    \includegraphics[width=0.32\textwidth]{figures/vstable.pdf}
	%\vspace*{-4ex}
    \caption{Win percentages of CIG 2012 competition, from left
    to right: 3-player maps only, 6-player maps only, all maps.
    Read from line to column, bot in row wins given fraction of
    games against bot in column. For some bots, we find interesting
    differences, e.g. Xelnaga gets worse on 6-player maps, UAlbertaBot
    gets better. Only Xelnaga can reliably beat Skynet, but only on 
    3-player maps}
    \label{fig:cig2012-mapresults}
\end{figure*}

@@

\subsection{StarCraft Bot Ladder}\label{sec:ladder}

@@

Real-time strategy games encompass many interesting and complex
sub-problems which are closely related not only to other fields of AI
research, but to real-world problems as well. For example, optimizing assembly line operations in factories is akin to performing build-order optimizations. 
Troop positioning in military conflicts involves the same spacial and
tactical reasoning used in RTS games. Robot navigation in unknown environments require real-time path-finding and decision making to avoid hitting obstacles.
All of these issues mentioned in this paper must are being tackled by
the real-time strategy game AI community, and in doing so we will not
only be improving techniques for writing tournament-winning bots, but
for advance the state of the art for many other fields as well.

%As argued before, addressing these open research questions would not just lead to better RTS game playing bots, but would provide means to improve many other complex AI components, not solely focussing on games. This would apply to many areas that are in need of real-time decision components working under uncertainty in the form of incomplete and/or noisy information, such as  e.g. robotics and assistance systems.
%bring significant advances to many other areas of artificial intelligence. This is so because RTS games embody many characteristics of many other problems appearing in many areas of AI (such as robotics, multiagent systems, etc.).
